Efficient pedigree recording for fast population genetics simulation

### 2017-12-09 forwards_paper.tex
merged jaime edits

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{peng2005simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, as implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a general-purpose method that uses these tools
(as implemented in \msprime{})
to efficiently record simplified a tree sequence,
and analyze its run time complexity.

@@

To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see below) with and without ancestry tracking.  
The simulations without ancestry tracking introduced neutral mutations at a rate 
equal to the recombination rate,
so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations and recombinations,
respectively, per gamete and per generation.
Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
We categorize simulations in terms of the \emph{scaled} parameters
$\theta = 4N\mu$ and $\rho = 4Nr$, where $N$ is the diploid population size,
since these most strongly determine run time.

We ran one simulation for each combination of $N$ from $(1000, 10000, 50000)$, and $\rho$ from $(10^3, 10^4, 10^5)$.
Deleterious mutations were introduced at rate $\rho/100$ per generation, 
drawing scaled selection coefficients ($2Ns$)
from a Gamma distribution with a mean of -5 and a shape parameter of 1.  
This distribution of fitness effects results in
thousands of weakly-deleterious mutations segregating in the population, 
many of which drift to intermediate frequencies.  
The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
coalescent \citep{Neuhauser1997-nn}.
Therefore, these selected mutations must be explicitly tracked in our forward simulation,
and the time savings due to ancestry tracking come from not having to record \textit{neutral} mutations.

@@

For efficiency reasons, we only ran \simupop{} with ancestry tracking.

\krt{Jaime--any details here to add re: how simuPOP was tweaked??}

\plr{Insert figures/tables of performance here and describe them.}

% \plr{Jaime and Kevin to describe simulation methods here (separately).}
% 
% Comparison of simulation with/without msprime, using \simupop{}
% or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
% 
% Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
% so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
% in a day.

>>>>

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{peng2005simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, as implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

To measure the performance gains from recording the pedigree we ran simulations with and
without recording, for both implementations (\simupop{} \citep{simupop}
and \fwdpp{} \citep{fwdpp}, as described below).
All simulations implemented a discrete-time Wright-Fisher population of diploid
population size $N$, simulated for $10N$ generations.
The simulations without pedigree recording introduced neutral mutations at a rate 
equal to the recombination rate,
so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations and recombinations,
respectively, per gamete and per generation.
% Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
% and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
We ran simulations with different values of $N$
and of the \emph{scaled} parameter $\rho = 4Nr$,
categorized by the value of $\rho$, since this
is what most strongly determines run time
(along with the scaled mutation rate, $\theta = 4N\mu$, which is also equal to $\rho$.)

Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
coefficients ($2Ns$)
from a Gamma distribution with a mean of -5 and a shape parameter of 1.  This distribution of fitness effects results in
thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
frequencies.  The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
coalescent \citep{Neuhauser1997-nn}.  Therefore, these selected mutations must be explicitly tracked in our forward simulation
and the time savings due to pedigree recording come from not having to record \textit{neutral} mutations.

@@

\subsubsection*{The \fwdpp{} implementation}

We implemented Wright-Fisher simulations in C++ using \fwdpp{} library
functions and interface code that records data into the \msprime{} Tables API (see below).
In these simulations, $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
\citep{Kimura1969-uc}.
We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

\plr{Perhaps implementations should go below in Methods?}
\plr{Insert figures/tables of performance here and describe them.}

\subsubsection*{The \simupop{} implementation}

We implemented Wright-Fisher simulations in Python using \simupop{} library
functions (which wrap C++ code) and interface code (\ftprime{}, also in Python) that records data into the \msprime{} Tables API (see below).
In these simulations, $\mu$ and $r$ define a per-base-pair mutation rate and we
defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

% Comparison of simulation with/without msprime, using \simupop{}
% or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
% 
% Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
% so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
% in a day.
